---
title: "Tutorial 2: Organise and manipulate data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


#  Organise data

## Vectors

*In the last tutorial, we encountered types of data: numbers (1,6, ...) and character strings ('blue', 'five', '5', 'fdbv 47.. df', ...). We organised them into vectors, that is a collection of elements with an order such that we can access for instance the third element of the vector.*

Below are two vectors, a numeric and  a character vector (the latter indicated by `'` or `"`). If you are not sure what type an R object is, you can type `class()`. In the box below, determine the type of each vector.

```{r vec-1, exercise=TRUE}

numeric_vector <- c(2,54,5,7)
character_vector <- c('two', 'flower','tree','2')

```


```{r vec-1-solution}
numeric_vector <- c(2,54,5,7)
character_vector <- c('two', 'flower','tree','2')

class(numeric_vector)
class(character_vector)
```

Next, output the third element of each vector using the `vector[3]` notation.

```{r vec-1b, exercise=TRUE}

numeric_vector <- c(2,54,5,7)
character_vector <- c('two', 'flower','tree','2')

```


```{r vec-1b-solution}

numeric_vector[3]
character_vector[3]
```


You can combine two vectors or add additional elements using again the function `c()`. Run the following code.

```{r vec-2, exercise=TRUE}

numeric_vector <- c(2,54,5,7)
character_vector <- c('two', 'flower','tree','2')

numeric_vector_2 <- c(numeric_vector, 3)
numeric_vector_2

c(numeric_vector, character_vector)
```

In the last line, we combined our character and numeric vector. R then automatically converted all numeric elements to character strings. The reason is that vectors can only contain one type of data. While numbers can be converted to character, the reverse is not always possible. 

(This holds true outside of R, for instance 5 can be the name of a room. Importantly, room 5 plus room 3 does not equal room 8. While this seems clear, it is still an often made mistake in statistics: Not everything that has a number should be treated as a number.)



## Data.frames

Typically, we need more than one vector to organise our data. For instance, we might want to study the height and age of a group of children. Our first vector contains the children's height measured in cm, that is the first element is the height of kid 1, 2, and so on. A second vector contains the age of child 1, 2, 3, etc. 

A straightforward way to organise these observations is to create at table in which the columns correspond to the different categories (variables), that is height or age. The rows correspond to the different observation units, in this case children. In addition, we probably want to add a third column with the name or identifier for each kid. The standard data structure for such a table in R is called `data.frame`. The box below defines and outputs a data frame.


```{r data-frame, exercise=TRUE}

name <- c('Peter', 'Helen', 'Ana', 'Sean', 'Chris', 'Mary')
age <- c(8,7,9,7,7,8)
height <- c(125,133,151,145,140,134)

age_height_df <- data.frame(name, age, height)

age_height_df
```

This brings us to a first look at the data we will use for our term project. I have added a subset of the Irish property data to this tutorial as a `data.frame`. The `data.frame` is called `irish_property_df`. Have a look at the data in the next box.  

```{r prop-data, exercise=TRUE}

irish_property_df
```

## Objects, vectors, ...

