---
title: "Tutorial 5: Graphs"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(quantsUL)
library(ggplot2)
library(stringr)
library(reshape2)
library(dplyr)

bairoch <- melt(Bairoch_1988, id.vars = c("country","city"))
bairoch$year <- as.numeric(str_extract(bairoch$variable, "[0-9]+"))
bairoch$population <- as.numeric(bairoch$value)*1000
bairoch$value <- NULL
bairoch$variable <- NULL

knitr::opts_chunk$set(echo = FALSE)
```


## ggplot2

***This tutorial is intended to start plotting in R using the package ggplots2. Please see Wickham et al's book 'ggplot2: Elegant Graphics for Data Analysis' for more background and tips (https://ggplot2-book.org). Use the 'R Graphics Cookbook' to explore different graphs quickly:  https://r-graphics.org/.***

***I have loaded the following packages into this tutorial: quantsUL, stringr, dplyr, and ggplot2.***

ggplot2 (principle is based on The Grammar of Graphics by Leland Wilkinson (1999), therefore the 'gg')  offers a systematic and comprehensive way to create common and novel graphs including histograms, scatter plots, or maps.

The basic principle is to iteratively define _layers_. For instance, we may first define the relevant variables, then the type of plot before fine tuning colours and labels.

I have added a dataset on population size of European cities based on "La Population Des Villes Europeennes" (Bairoch, Batou, Chevre 1988). The name of the data is `bairoch`. 

Press run to show the first observations (rows). 

Can you change the code to display only observations from Ireland?


```{r Daten, exercise=TRUE}

head(bairoch)

```
```{r Daten-solution}

# only Ireland
head(bairoch[bairoch$country=="Ireland",])

# dplyr alternative
bairoch |> 
  filter(country=="Ireland") |> 
  head()

```

In addition, can you change the code to display only observations from Ireland without rows that have no information (`NA` for population)?

*Hint: `is.na(bairoch$population)` creates a vector that is TRUE if population is NA and FALSE otherwise.*

```{r Daten2, exercise=TRUE}

head(bairoch)

```
```{r Daten2-solution}

# only Ireland and population is not NA
head(bairoch[bairoch$country=="Ireland" & !is.na(bairoch$population),]) 

# dplyr alternative
bairoch |> 
  filter(country=="Ireland" & !is.na(bairoch$population)) |> 
  head()

```

We will now plot the population of Limerick over time. We start with the function `ggplot()`to create the basic ggplot layer. Next we add a layer for points that represent years on the x-axis and population on the y axis using the  function `geom_point()`. The function `aes` is a helper function that maps the variables to visual properties, in this instance the position along the x and y-axis. We will look a this function later again.

Multiple layers are combined with the plus sign.

```{r Dortmund, exercise=TRUE}
# subset the data to complete observations for Limerick
Limerick <- bairoch |> filter(city=="Limerick" & !is.na(bairoch$population)) 

# create plot as R object
LimerickPlot <- ggplot(data = Limerick) + # base layer
  geom_point(mapping = aes(x=year, y=population)) # add points for year and population

# output plot
LimerickPlot
```

Now ,we want to add another layer to our plot to connect the dots with lines. Since we will use the same x and y axis for points and for lines, we move the `aes()` function to the first, base layer.

```{r Dortmund2, exercise=TRUE}
# create data subset
Limerick <- bairoch |> filter(city=="Limerick" & !is.na(bairoch$population)) 

# create plot and output directly 
ggplot(data = Limerick, mapping = aes(x=year, y=population)) +
  geom_point() +
  geom_line()

```


## Groups

We want to plot the data for multiple cities in the same plot. To keep the observations separate, we have to tell data what rows belongs to what group (city). Using again the `aes` function, we create groups from the data based on the variable `city`.

```{r group, exercise=TRUE}
# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# group by city
cityPlot <- ggplot(data=Limerick, mapping = aes(x=year, y=population, group=city)) + 
  geom_point() +
  geom_line() 

cityPlot
```

To distinguish the observation points and lines, we need to add more visual properties such as colour or shape using the `aes()` function. Once we use colour, R knows how to group the observations so we can leave out the group argument.

```{r color, exercise=TRUE}

# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# replace group with colour in aes function
cityPlot <- ggplot(data=Limerick, mapping=aes(x=year, y=population, colour=city)) +
  geom_point() +
  geom_line() 

cityPlot

```

We can also use more than one visual property to distinguish groups. Visual properties that are not differentiated by variable can be placed outside the `aes()` function

```{r color2, exercise=TRUE}

# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# change point shape by city
ggplot(data=Limerick, mapping=aes(x=year, y=population, colour=city)) +
  geom_point(aes(shape=city)) +
  geom_line() 

# increase dot size 
ggplot(data=Limerick, mapping=aes(x=year, y=population, colour=city)) +
  geom_point(mapping=aes(shape=city), size=3) +
  geom_line() 

```

An alternative approach is to use the layer `facet_wrap()` to create separate plots by group.

```{r facet, exercise=TRUE}

# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# use facet_wrap to create plot by one or two groups (2nd group goes before ~)
ggplot(data=Limerick, mapping=aes(x=year, y=population)) +
  geom_point() +
  geom_line() +
  facet_wrap(facets = ~ city)

```

It depends on the plot type, data, and our focus what method to plot multiple groups is preferable.


## Graphical parameters

After deciding on a graph type, we can add more layers to improve clarity and aesthetics. For instance, we can relabel the axis and move the legend to the bottom

```{r labs, exercise=TRUE}

# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# add lab titles and move the legend
ggplot(Limerick, aes(x=year, y=population, colour=city)) +
  geom_point(aes(shape=city)) +
  geom_line(aes(linetype=city)) +
  labs(x="Year", y="Population", colour="", shape="", linetype="") +
  theme(legend.position = "bottom")


```


You will find a huge variety of options online or in the afore mentioned book https://r-graphics.org/. For instance, you can use the same style as the BBC: https://bbc.github.io/rcookbook/.

The package `viridis` offers colour combination that help, ofr instance, to differentiate colours for people with colour blindness or if printed black and white. 

We will also use a black and white theme instead of the standard ggplot colour layout (`theme_bw`).

```{r vir, exercise=TRUE}

library(viridis)
# create data subset
Limerick <- bairoch |> filter(city %in% c("Limerick", "Cork", "Galway") & !is.na(bairoch$population)) 

# add theme_bw() and scale_color_viridis_d, note that theme() needs to be called after theme_bw
ggplot(Limerick, aes(x=year, y=population, colour=city)) +
  geom_point(aes(shape=city)) +
  geom_line(aes(linetype=city)) +
  labs(x="Year", y="Population", colour="", shape="", linetype="") +
  theme_bw() +
  scale_color_viridis_d() +
  theme(legend.position = "bottom") 


```

As a final exercise, we will change the type of plot. Importantly, the principles introduced so far apply almost all types of plots, for instance histograms or barplots. Explore the options at https://r-graphics.org/ and online, for instance via Google Images.

```{r hist, exercise=TRUE}

ggplot(bairoch, aes(population)) +
    geom_histogram()


```

The difficult is different to read. Can you improve the plot? You can see the parameters in the help file (`?geom_histogram`)? Or would you rather use a different type of plot?


## Dealing with data problems

The main difficulty (beside picking the plot type) often lies in the data preparation. `ggplot` requires the data to be in a certain format. A common task is to summarise data before plotting. The functions `aggregate` (base R) or `summarise` (dplyr) can help.

We had seen the subset of rows for Ireland in the beginning of this tutorial. 

```{r agg0, exercise=TRUE}
bairoch |> 
  filter(country=="Ireland" & !is.na(population)) 
```

Can you sum up the (urban) population by year similar to the last tutorial?

*Data caveat: All cities have years without population information, the aggregates are not the full urban population.*

```{r agg, exercise=TRUE}


```
```{r agg-solution}

# base R
aggIrl_1 <- aggregate(population ~ year, 
                      data=bairoch[bairoch$country=="Ireland" & !is.na(bairoch$population),], 
                      FUN=sum)
aggIrl_1

# dplyr
aggIrl_2 <- bairoch |> 
  filter(country=="Ireland" & !is.na(population)) |> 
  group_by(year) |> 
  summarise(population = sum(population))
aggIrl_2
```

We can then plot the aggregates over time. 

```{r aggPlot, exercise=TRUE}
aggIrl <- bairoch |> 
  filter(country=="Ireland" & !is.na(population)) |> 
  group_by(year) |> 
  summarise(population = sum(population))

ggplot(aggIrl, aes(x=year, y=population)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(y="urban population")

```

We can use the same approach to aggregate over more variables, for instance year and country.

```{r agg2, exercise=TRUE}

# base R
aggIrl_1 <- aggregate(population ~ year + country, 
                      data=bairoch[bairoch$country %in% c("Ireland", "Portugal") & !is.na(bairoch$population),], 
                      FUN=sum)
aggIrl_1

# dplyr
aggIrl_2 <- bairoch |> 
  filter(country %in% c("Ireland", "Portugal") & !is.na(population)) |> 
  group_by(year, country) |> 
  summarise(population = sum(population))
aggIrl_2
```

Often, it is easier to restrict the sample later, for instance, if we want to explore multiple country combinations. 

Try different combinations of countries in the code below and check the effect of changing, for instance, the labels or theme.

```{r aggPlot2, exercise=TRUE}
# aggregate
aggIrl <- bairoch |> 
  filter(!is.na(population)) |> 
  group_by(country, year) |> 
  summarise(population = sum(population))

# print name of all countries in data
unique(aggIrl$country)

# plot population over time by countries for comparison 

aggIrl |> 
  filter(country %in% c("France", "Spain", "Great Britain")) |> 
  ggplot(aes(x=year, y=population, colour=country)) +
    geom_point(aes(shape=country)) +
    geom_line() +
    scale_colour_viridis_d() +
    theme_bw() +
    labs(y="urban population", colour="", shape="")

```

